\chapter{Background and Literature Review}
\label{chap:background}

This chapter reviews the literature along the lines most relevant to the comparative evaluation of Chapters~\ref{chap:methods}--\ref{chap:results}: the motivation for soft-robot control and why the inverse is hard, the modelling and learning method families, sensing and simulation infrastructure, evaluation baselines, and the research gap the thesis is positioned against.

\section{Soft robots and the control motivation}
\label{sec:bg_motivation}

Soft and continuum robots are attractive in human-facing and confined-space environments because they can bend continuously, conform to irregular geometries, and interact more safely with delicate surroundings than rigid-link manipulators \cite{rus2015softrobots, burgnerkahrs2015medical, dupont2022medical}. This thesis focuses on the pneumatic subclass: silicone cylinders with several internal chambers that bend and elongate under pressurised air.

The same compliance that makes these arms safe also makes them difficult to control: the body deforms continuously under distributed pressure rather than moving through a small set of rigid joints, so the map from actuation to tip position is nonlinear, design-specific, and history-dependent through viscoelasticity, pneumatic lag, and fabrication variability \cite{rus2015softrobots, thuruthel2018survey, kim2021mlsoft}. The control problem can be written as
\begin{equation}
  \mathbf{x}_{t} = f(\mathbf{u}_{t};\, \boldsymbol{\theta}, \boldsymbol{\eta}_{t}),
  \label{eq:bg_forward}
\end{equation}
with $\mathbf{u}_{t}$ the actuation command, $\mathbf{x}_{t}$ the tip position, $\boldsymbol{\theta}$ the robot-dependent parameters, and $\boldsymbol{\eta}_{t}$ the latent history; the inverse problem recovers a $\mathbf{u}$ that attains a target $\mathbf{x}^{\ast}$.

The pneumatic platform of Dataset~1 is the I-Support manipulator \cite{manti2016soft, ansari2017assistive, zlatintsi2020isupport}; the version used here is the additively-manufactured AM-I-Support, in which the original latex--silicone chambers are replaced by 3D-printed thermoplastic polyurethane of Shore~80A hardness \cite{centurelli2022closed, alessi2024pushing}. Two modules are assembled in series with a $60^{\circ}$ inter-module offset, giving six chambers covering the full actuation circle \cite{bianchi2024softoss}. Dataset~2 is an undocumented three-valve recording of unknown geometry and material, treated as a black box throughout. Detailed parameters of both datasets are given in Section~\ref{sec:datasets}; this chapter refers to them only to fix the regime---over-actuated six-chamber versus under-documented three-valve---against which the literature claims are later tested.

\section{Why inverse kinematics of pneumatic soft robots is hard}
\label{sec:bg_ik_hard}

Soft-robot inverse kinematics can be formalised as a constrained optimisation,
\begin{equation}
  \mathbf{u}^{\ast} = \arg\min_{\mathbf{u}}\; \| f(\mathbf{u};\, \boldsymbol{\theta}, \boldsymbol{\eta}) - \mathbf{x}^{\ast} \|_{2}^{2} + \lambda\, R(\mathbf{u}),
  \label{eq:bg_ik_obj}
\end{equation}
where $R(\mathbf{u})$ encodes feasibility, smoothness, energy, or branch preference. Three structural challenges make the inverse harder than the forward map.

\paragraph{Set-valued inverse.}
When the actuation dimension $n_{u}$ exceeds the task-space dimension $n_{x}$, several commands place the tip at the same point with different internal shapes. A direct regressor trained with squared error is then asked to approximate a single branch of a relation, and can collapse to an averaged command that is physically suboptimal or unrealisable \cite{wang2021mlsurvey, giorelli2015nn, fang2022jacobian}. Dataset~1 (six chambers, three-dimensional tip) is the canonical over-actuated case.

\paragraph{History dependence.}
Soft pneumatic actuators show rate effects, creep, hysteresis, and flow dynamics, so the same target may require different inputs depending on the recent trajectory of the state \cite{thuruthel2018survey, thuruthel2019modelbased, tariverdi2021rnn}. A static regressor will miss this dependence unless the observable input already carries information about the latent state.

\paragraph{Partial observability.}
End-point sensing alone typically under-identifies the internal configuration of a deformable body; partial-shape observations (a mid-body marker, strain sensor, or proprioceptive signal) change the identifiability structure, not just the input features \cite{sincak2024sensing, wang2024proprioception}. The tip-plus-mid ablation reported in Chapter~\ref{chap:results} tests this empirically.

\section{Modelling hierarchy for soft-robot control}
\label{sec:bg_models_hierarchy}

The literature organises along an analytical-tractability versus physical-fidelity trade-off summarised in Table~\ref{tab:bg_models} \cite{webster2010review, rucker2011statics, gazzola2018pyelastica, mengaldo2022embodied, armanini2023structured}.

\paragraph{Piecewise constant curvature (PCC).}
\label{sec:bg_models_pcc}
PCC approximates each segment as a circular arc of constant curvature and supports closed-form or near-closed-form inverse kinematics \cite{webster2010review}; it is the natural analytical baseline for many soft-robot studies and is used as such in Sections~\ref{sec:pcc_single}--\ref{sec:pcc_multi}. The constant-curvature assumption fails under gravity, torsion, asymmetric reinforcement, extension, or contact \cite{webster2010review, mengaldo2022embodied, armanini2023structured}; when those effects are substantial the model is not merely noisy but biased by design.

\paragraph{Cosserat-rod models.}
\label{sec:bg_models_cosserat}
Rod theory treats the backbone as a slender elastic body with centreline and cross-section orientation varying continuously along arclength, capturing nonconstant curvature, extension, shear, twist, and distributed loads \cite{rucker2011statics, gazzola2018pyelastica, mengaldo2022embodied, armanini2023structured}. The PyElastica library \cite{gazzola2018pyelastica} is the simulator adopted in Section~\ref{sec:sim_setup}.

\paragraph{Finite-element modelling (FEM).}
Volumetric discretisation captures complex material laws, geometric nonlinearities, and contact at high computational cost and heavy parameter identification \cite{mengaldo2022embodied, armanini2023structured}; FEM is rarely inverted at run time and is instead used for offline data generation or as a physics prior \cite{gao2024residual, park2025dualae}.

\paragraph{Implication for this thesis.}
PCC is retained as the analytical baseline (what a tractable geometric approximation can and cannot recover); a Cosserat rod is the sim-to-real source; learned surrogates sit at every level.

\begin{table}[ht]
\centering
\caption[Modelling levels used in soft-robot control.]{Modelling levels used in soft-robot control and the role each plays in this thesis.}
\label{tab:bg_models}
\begin{tabularx}{\linewidth}{@{}l X X X X@{}}
\toprule
Modelling level        & Main state description                         & Strengths                                              & Limitations                                                  & Role in this thesis \\
\midrule
PCC (single segment)   & Circular-arc section $(\kappa, \varphi, L)$     & Fast, interpretable, closed-form intuition             & Breaks under nonconstant curvature, extension, gravity        & Analytical baseline (Section~\ref{sec:pcc_single}) \\
Multi-segment PCC      & Chained PCC sections                            & Preserves tractability for modular arms                 & Inverse becomes redundant and branch-sensitive                & Reference model for over-actuated arms (Section~\ref{sec:pcc_multi}) \\
Cosserat/rod models    & Continuous backbone strain and pose fields      & Captures variable curvature, extension, shear, twist    & Slower and more calibration-intensive than PCC                & Physics prior / simulator (Section~\ref{sec:sim_setup}) \\
FEM                    & Volumetric deformation                          & Highest physical detail                                 & High cost and difficult identification                        & Offline digital twin / synthetic data \\
Learned surrogate      & Data-driven mapping in input--output space      & Fast inference; absorbs mismatch                        & Depends on data coverage and observability                    & Inverse model / learned forward arbiter (Sections~\ref{sec:mlp}, \ref{sec:jacobian}) \\
\bottomrule
\end{tabularx}
\end{table}

\section{Machine learning and inverse-kinematics method families}
\label{sec:bg_ml_ik}

The relevant taxonomy, summarised in Table~\ref{tab:bg_methods}, distinguishes direct inverse learning, temporal or dynamic learning, and hybrid learned-forward inversion, with reinforcement learning as an adjacent strand \cite{thuruthel2018survey, kim2021mlsoft, wang2021mlsurvey}.

\paragraph{Direct inverse learning.}
\label{sec:bg_ik_direct}
A feed-forward network $\hat{\mathbf{u}} = g_{\psi}(\mathbf{x}^{\ast})$ absorbs fabrication imperfections and nonlinearities directly from data \cite{kim2021mlsoft, wang2021mlsurvey}; Giorelli et al.\ \cite{giorelli2015nn} and Ma et al.\ \cite{ma2022bp} showed it outperforms analytical baselines on soft arms. The weakness is structural: on a multi-valued inverse the learned output drifts toward the branch-average \cite{wang2021mlsurvey, fang2022jacobian}, the failure mode that Dataset~1's feed-forward baseline exhibits in Chapter~\ref{chap:results}.

\paragraph{Temporal / recurrent models.}
\label{sec:bg_ik_temporal}
Recurrent architectures introduce a latent state that carries history, capturing hysteresis and flow delay statistically rather than through explicit constitutive models \cite{thuruthel2019modelbased, tariverdi2021rnn}. Temporal context helps only if the past carries information about the hidden state: for i.i.d.\ data or impoverished features, recurrent models add complexity without solving identifiability \cite{wang2021mlsurvey, tariverdi2021rnn}.

\paragraph{Jacobian-learned hybrid.}
\label{sec:bg_ik_jac}
One learns the forward map $f_{\theta}(\mathbf{u})$ and inverts it iteratively via damped least-squares:
\begin{equation}
  \mathbf{u}_{k+1}
  = \mathbf{u}_{k}
    + J_{f}(\mathbf{u}_{k})^{\top} \left( J_{f}(\mathbf{u}_{k}) J_{f}(\mathbf{u}_{k})^{\top} + \lambda^{2} I \right)^{-1} \big( \mathbf{x}^{\ast} - f_{\theta}(\mathbf{u}_{k}) \big).
  \label{eq:bg_jacobian_update}
\end{equation}
The forward map is single-valued and therefore easier to learn, and the iteration commits to a single branch rather than averaging \cite{fang2022jacobian}. The limitation is that performance is capped by the learned forward model.

\paragraph{Reinforcement learning.}
Policy-learning work \cite{thuruthel2019modelbased} extends soft-robot control beyond static inverse kinematics but is treated here as adjacent context rather than a head-to-head comparator, since this thesis holds dataset, metric, and protocol fixed across IK families.

\begin{table}[ht]
\centering
\caption[Inverse-kinematics method families relevant to this thesis.]{Inverse-kinematics method families relevant to this thesis.}
\label{tab:bg_methods}
\begin{tabularx}{\linewidth}{@{}l X X X X@{}}
\toprule
Method family                    & Core idea                                       & Main advantage                                             & Main failure mode                                         & Most suitable regime \\
\midrule
Analytical PCC IK                & Invert a simplified geometric model              & Fast, interpretable, no training data                      & Structural bias when curvature/extension assumptions fail  & Simple, well-characterised, low-redundancy platforms \\
Feed-forward neural IK           & Learn $\mathbf{x}^{\ast} \mapsto \mathbf{u}$ directly & Strong empirical baseline, robust to mismatch           & Branch averaging on multi-valued inverses                 & Single-segment or well-observed platforms \\
Recurrent / temporal IK          & Learn inverse with a short history window        & Can exploit hysteresis and dynamic context                 & Helps little when history is uninformative                & History-dependent platforms with meaningful temporal structure \\
Jacobian-learned IK              & Learn forward and invert iteratively             & Uses easier-to-learn forward map; avoids branch collapse   & Depends on FK quality and initialisation                   & Redundant platforms where position accuracy matters most \\
Sim-pretrained / hybrid transfer & Simulation prior, fine-tune on real data         & Reduces real-data burden when prior is structurally correct & Negative transfer under simulator--target mismatch        & Platforms with good simulator--target overlap \\
\bottomrule
\end{tabularx}
\end{table}

\section{Sensing, proprioception, and observability}
\label{sec:bg_sensing}

Control quality depends on what is observed: end-effector sensing alone often under-identifies the internal configuration of a deformable body. Sincak et al.\ \cite{sincak2024sensing} reframe control performance as a sensing problem; Wang et al.\ \cite{wang2024proprioception} show that a mid-body observation is not just another input feature but an intervention that changes the identifiability of the problem.

For Dataset~1, the two-module AM-I-Support arm is driven by six actuation channels while only the tip is observed in the minimal sensing case, so the learning system is asked to infer a six-dimensional command from a three-dimensional target. A mid-segment observation makes the inverse better constrained; the tip-only versus tip-plus-mid comparison reported in Chapter~\ref{chap:results} tests this observability hypothesis.

\section{Simulation and sim-to-real transfer}
\label{sec:bg_sim}

Simulation provides synthetic training data, safe controller prototyping, and a structured prior for transfer learning \cite{gazzola2018pyelastica, fang2022jacobian, peng2018simtoreal, gao2024residual, park2025dualae}. The simulator used in this thesis is PyElastica \cite{gazzola2018pyelastica}, a Python implementation of the Cosserat-rod formulation above.

Four sim-to-real strategies have emerged. \emph{Direct transfer} uses simulator weights unchanged on the real robot and is typically a lower baseline. \emph{Fine-tuning} a sim-pre-trained network on a small real-data budget is the most common soft-robotics route \cite{fang2022jacobian}. \emph{Domain randomisation} samples simulator parameters from distributions that span the expected real range \cite{peng2018simtoreal}; it is harder to apply to soft robots whose gaps include hysteresis, viscoelastic creep, and manufacturing asymmetries. \emph{Real-to-sim calibration and residual physics}, as in Gao et al.\ \cite{gao2024residual} and Park et al.\ \cite{park2025dualae}, either tune the simulator against real data or learn a residual correction on top of it. Across the four, the recurring finding is that transfer value is bounded above by geometric and kinematic overlap between simulator and real robot---the finding the PyElastica experiments of Chapter~\ref{chap:results} reproduce on Dataset~2.

\section{Evaluation metrics and published baselines}
\label{sec:bg_metrics}

Position-space error (Euclidean tip distance, optionally normalised by arm length) and pressure-space error (MAE, MAPE, NMAE, $R^{2}$) are the two standard metric families; trajectory-tracking evaluations additionally report command smoothness and inference time. Three published results serve as scale markers: Giorelli et al.\ \cite{giorelli2015nn} reported $7.35$\,mm mean tip error ($2.62\%$ of a $280$\,mm cable-driven arm) for a feed-forward network; Ma et al.\ \cite{ma2022bp} reported pressure-space MAPE below $5\%$ for a BP network on a $216$-sample three-chamber pneumatic arm; Fang et al.\ \cite{fang2022jacobian} reported sub-millimetre closed-loop error for a Jacobian-learned iteration on a pneumatic arm with a digital-twin simulator. The three bracket the accuracy range cleanly-characterised soft-robot platforms have attained.

\section{Research gap}
\label{sec:bg_gap}

Three observations about the surveyed literature set up the gap the thesis addresses.

\paragraph{Gap 1: method families are rarely compared head-to-head under matched protocols.}
PCC, feed-forward, recurrent, and Jacobian-learned methods have been reported on different platforms, with different sensing assumptions and different evaluation metrics \cite{webster2010review, thuruthel2018survey, kim2021mlsoft, wang2021mlsurvey, giorelli2015nn, ma2022bp, thuruthel2019modelbased, tariverdi2021rnn, fang2022jacobian}. Practitioners choosing between them must assemble non-equivalent experiments across papers.

\paragraph{Gap 2: partial-shape sensing is recognised but insufficiently isolated.}
The sensing literature makes clear that internal-shape information matters \cite{sincak2024sensing, wang2024proprioception}, yet controlled ablations that change only the observation while keeping the pipeline fixed are rare.

\paragraph{Gap 3: sim-to-real under mismatched priors is under-studied.}
Positive transfer results tend to come from platforms aligned with the simulator \cite{fang2022jacobian, gao2024residual, park2025dualae}; the common case of borrowing a simulator from a related platform for an under-documented target has received less systematic attention.

\paragraph{Thesis claim.}
A concise claim emerging from the literature and tested in Chapter~\ref{chap:results} is that \emph{inverse-kinematics method choice for pneumatic soft robots is regime-dependent rather than universal, and the regime is set primarily by actuation redundancy, sensing coverage, and simulator--target overlap}. The contributions of Chapter~\ref{chap:introduction} are organised to validate each part of this claim in turn.
