\chapter{Materials and Methods}
\label{chap:methods}

% TODO: bridging paragraph outlining the experimental programme (Exp 0--6
%       from the project plan).

\section{Datasets}
\label{sec:datasets}

Two pre-recorded datasets were used throughout this project. The first (DS1) was released with the I-Support soft manipulator developed by Manti et al.\ \cite{manti2016soft} and extended to a two-module configuration by Bianchi et al.\ \cite{bianchi2023learning}. The second (DS2) was distributed as a single file \texttt{all\_data\_R01.csv} [Insert citation for DS2 source once confirmed]. Both datasets contain pressure or valve commands paired with optical marker measurements, and therefore provide the input--output pairs required for learning-based inverse kinematics (IK). No hardware was operated in this study; all analyses and experiments were conducted on the archived data.

\subsection{DS1 --- I-Support (20\,Hz)}
\label{sec:ds1}

The I-Support platform is a silicone-based soft manipulator composed of two identical modules connected in series. Each module is 205\,mm long at rest, 60\,mm in diameter, and contains three McKibben-type fluidic chambers distributed uniformly at $120^{\circ}$ around the module axis. The second module is rotated by $60^{\circ}$ relative to the first, so that the six chambers together cover the full actuation circle. The robot was mounted vertically with the tip pointing downwards, giving a rest length of approximately 410\,mm and minimising the gravitational load on the end-effector. The maximum safe chamber pressure is 1.2\,bar (120\,kPa), and the pressure--deformation response is nonlinear, with a dead zone up to approximately 0.5\,bar \cite{manti2016soft}.

Nine reflective markers were attached to the robot in three triads, placed on the fixed base, on the mid-section between the two modules, and on the end-effector. A random-walk pressure sequence was generated and delivered to the six chambers at a sampling rate of 20\,Hz. Five recordings of 4{,}000 samples each were provided, corresponding to four NumPy seeds (0, 1, 2, 3). Seed~0 was repeated three times (\texttt{s0t1}, \texttt{s0t2}, \texttt{s0t3}) in order to quantify run-to-run repeatability. Each recording is stored as a single array of shape $(4000, 33)$: the first 27 columns hold the concatenated marker coordinates in millimetres, and the last six hold the chamber pressures in raw units where a value of approximately 100 corresponds to 1\,bar. Marker occlusions were present in roughly 0.5\% of the raw frames, and mid- and tip-marker occlusions were found to co-occur, so that the same subset of frames remained usable for every input variant considered in this work.

\subsection{DS2 --- R01 pneumatic recording}
\label{sec:ds2}

DS2 is distributed as a single comma-separated file of 18{,}072 rows with six columns: three valve commands (\texttt{V1}, \texttt{V2}, \texttt{V3}) and the tip position (\texttt{X}, \texttt{Y}, \texttt{Z}). The valve commands are integer-valued within the range $[0, 2500]$; the mapping to physical pressure is not documented in the release, and the commands were therefore treated as an un-calibrated actuation coordinate throughout this work [Insert pressure calibration information from the original source if it becomes available]. Tip positions were supplied in inches and were converted to millimetres by multiplying by 25.4 before any further processing.

After conversion, the tip coordinates spanned approximately $[-47.9, 53.8]$\,mm in $X$, $[-47.7, 56.1]$\,mm in $Y$, and $[-27.5, 36.9]$\,mm in $Z$, describing a small single-segment workspace roughly 100\,mm wide and 60\,mm deep. The file contains no missing values and no duplicate rows. The rows are ordered as recorded and are treated here as a single sequence; the presence or absence of temporal structure was examined in the exploratory analysis (Section~\ref{sec:eda}) and informed the interpretation of the sequence models in Experiment~3, but the order itself was preserved in every preprocessing step. The source robot for DS2 is a single-segment three-valve pneumatic manipulator [Insert manufacturer, geometry, and citation when confirmed]; because its physical parameters were not included in the release, DS2 was treated as a black-box actuation-to-pose mapping, and geometric quantities were never assumed without verification.

\subsection{Data splits and preprocessing}
\label{sec:splits}

DS1 was split at the seed level. Seeds~1 and~2 (\texttt{s1t1}, \texttt{s2t1}) formed the training set, seed~3 (\texttt{s3t1}) served as the validation set, and the three repetitions of seed~0 constituted the test set. This arrangement keeps all three repeated runs together, so that intra-seed repeatability can be reported as a variance estimate on the test metric and used as a physical noise floor for the IK error [Insert measured seed-0 repeatability from Section~\ref{sec:eda}]. Samples from the same seed were never mixed across splits. DS2 was split chronologically in the ratio $70/15/15$, yielding 12{,}650 training, 2{,}710 validation, and 2{,}712 test frames. Random shuffling was not applied at any stage, in order to preserve the original recording order and to avoid leaking information between splits should any residual temporal structure be present.

Tip and mid coordinates in DS1 were expressed relative to the base triad on a per-frame basis. For each frame, the base position was computed as the mean of the three base markers, and the same per-frame subtraction was applied to the mid and tip means. This operation removes any static translation offset introduced by re-installing the robot between recordings, and is sufficient so long as no rotational drift occurs between sessions; the assumption was checked in the exploratory analysis (Section~\ref{sec:eda}) by comparing the per-seed workspace point clouds. Frames in which either the mid or the tip triad contained a missing marker were dropped jointly, leaving 23{,}936 usable frames across the five recordings. Dropping frames jointly, rather than separately per input variant, ensured that the tip-only and tip-plus-mid configurations of Experiment~2 could be compared on an identical sample set.

Inputs and outputs were standardised to zero mean and unit variance using \texttt{StandardScaler} fitted on the training split only. Input and output scalers were fitted independently, and both were stored inside the model checkpoint metadata so that inference could be reproduced without recomputing dataset statistics. For the sequence models in Experiment~3, standardisation was applied before the sliding-window construction. Sliding windows were constrained to remain within a single seed in DS1 and within a single split in DS2, so that no window crossed a recording boundary or a split boundary.

A fixed random seed of 42 was used for all stochastic operations, including scaler fitting order, mini-batch shuffling within a single DS1 seed, and weight initialisation.

\section{Exploratory Data Analysis (Exp 0)}
% TODO: workspace shape, marker-occlusion rate, time-series verification
%       for DS2, pressure--tip correlation.

\section{Analytical Baseline: PCC Inverse Kinematics}

\subsection{Single-Segment PCC (Exp 1A, DS2)}
% TODO: equations from Ma et al. (2022) eq. 11--21 (see
%       docs/references/pcc_equations.md).

\subsection{Two-Segment PCC (Exp 1B, DS1 --- stretch goal)}
% TODO: compound PCC + SQP optimisation (see
%       docs/references/multi_segment_pcc.md).

\section{Feed-Forward Neural IK}

\subsection{MLP Architecture (Exp 2A, Tip-Only)}
% TODO: inputs (X,Y,Z), hidden layers, activation, outputs = pressures.

\subsection{Tip + Mid Ablation (Exp 2B, DS1)}
% TODO: augment inputs with mid-segment marker mean; motivation: the
%       two-segment arm is underdetermined from tip alone.

\section{Temporal IK with LSTM (Exp 3)}
% TODO: windowing strategy, sequence length, stateful vs. stateless,
%       expected benefit on hysteresis.

\section{Jacobian-Learned IK (Exp 4)}
% TODO: learned-Jacobian + damped least-squares update rule from
%       Fang et al. (2022); see docs/references/jacobian_ik_algo.md.

\section{Simulation and Sim-to-Real with PyElastica (Exp 5)}

\subsection{Simulator Setup}
% TODO: Cosserat rod params from Manti 2016; pneumatic actuation model.

\subsection{Sim-to-Real Transfer Protocol}
% TODO: pre-train on synthetic, fine-tune on real DS2; evaluation protocol.

\section{Trajectory Tracking Evaluation (Exp 6)}
% TODO: unified protocol (figure-of-eight, circle, random) applied to
%       every IK method.

\section{Training Setup and Evaluation}

\subsection{Hardware, Software and Reproducibility}
% TODO: GPU, PyTorch version, seed = 42, checkpoint metadata layout.

\subsection{Hyperparameter Selection}
% TODO: validation-set driven search; never touch the test set.

\subsection{Metrics}
% TODO: MAE / RMSE / MAPE of tip position, trajectory error, inference
%       latency.
