\chapter{Discussion}
\label{chap:discussion}

\section{Interpretation of results}
\label{sec:disc_interpretation}

No single method dominated every regime: each won on some dataset and lost elsewhere. This section reads the results as one story, argues that the regimes depend on three physical properties, and explains the least obvious findings.

\paragraph{Classical baselines failed for different reasons.}
Single-segment PCC hit a $22$\,mm residual on Dataset~2's $\sim$$100$\,mm workspace because an elongation signal ($r = +0.737$ with total pressurisation) dominates and a constant-$L$ arc cannot represent it. Two-segment PCC reached a geometric residual of $3 \times 10^{-3}$\,mm on Dataset~1 but a $51$\,mm closed-loop error: the multi-valued inverse lands on different solution branches across training frames, and ordinary least-squares calibration collapses to the per-chamber mean. Correct geometry therefore does not guarantee a competitive inverse---both configuration-space and actuation-space well-posedness are required.

\paragraph{Data-driven regression tolerates ill-posedness by averaging.}
A squared-loss regressor maps a multi-valued inverse to the per-tip posterior mean over valid pressures. Dataset~1's tip-only network plateaued at $R^{2} = 0.32$ after four epochs (under-determined); Dataset~2's single-valued inverse reached $R^{2} = 0.93$ with the same architecture. The predictive ceiling is set by the posterior's concentration, not by the network's capacity: the right diagnostic is actuation redundancy, not architecture choice.

\paragraph{Mid-marker lifts the ceiling; temporal context only helps once it does.}
Adding the mid-marker was the single largest improvement: $R^{2}$ rose from $0.32$ to $0.44$ (MLP) and from $0.28$ to $0.51$ (LSTM). The tip-only LSTM was marginally worse than the tip-only MLP, so temporal context is uninformative without observable internal state; combined with the mid marker, the recurrent network extracts the internal-state trajectory, which is physically meaningful only once the state itself is observable.

\paragraph{Position vs pressure trade-off is an inverse-problem property.}
On multi-valued inverses a direct regressor minimises pressure error (posterior mean); Jacobian iteration commits to a single branch and minimises position error. Dataset~2 showed a $33\%$ pressure-MAE increase for a $71\%$ position-error reduction (Table~\ref{tab:res_jacobian_pos}); Dataset~1 similar in sign, smaller in magnitude. The two methods therefore measure different objectives---pick by downstream use.

\paragraph{Sim-to-real dominated by workspace geometry, not sample count.}
Strategy~B matched Strategy~A at every budget; Strategy~C reached $R^{2} = -1.71$. The synthetic cloud was $5\times$ Dataset~2's extent on each axis and differently centred, so the prior pointed at an irrelevant region; fine-tuning on real data simply discarded it. A Manti-inspired rod is not a drop-in prior for an undocumented robot---the $720$-sample count was not the ceiling.

\paragraph{A compact decision criterion.}
The regime is set by three arm properties. (1) \emph{Actuation redundancy}: when DOF exceed tip-DOF, Jacobian iteration reaches the tolerance floor. (2) \emph{Sensing coverage}: when the tip under-identifies the internal state, one partial shape observation beats any network-side intervention. (3) \emph{Target-domain overlap}: a simulator prior is useful only when its synthetic workspace covers the real operating region. These three account for every regime-specific finding in Chapter~\ref{chap:results}.

\section{Comparison with published baselines}
\label{sec:disc_benchmarks}

This section positions the headline numerical findings of Chapter~\ref{chap:results} against three reference points in the soft-robotics inverse-kinematics literature: Ma et al.\ \cite{ma2022bp} on a purpose-built three-chamber pneumatic arm, Giorelli et al.\ \cite{giorelli2015nn} on a cable-driven soft arm, and Fang et al.\ \cite{fang2022jacobian} on a pneumatic platform with sim-to-real learning. Each reference uses a different robot, dataset, and metric, so the comparisons below are organised to separate what can be compared directly from what requires cross-platform caveats. Table~\ref{tab:disc_benchmarks} collects the position-space tracking numbers for the references and for the two best configurations of this thesis, normalised by the rest length of each arm so that absolute millimetres are not the only scale.

\begin{table}[ht]
\centering
\caption[Position-space tracking error against published baselines.]{Position-space tracking error for three published reference points and the two best configurations of this thesis, normalised by the rest length of the corresponding arm. The $280$\,mm length is taken from Giorelli et al.\ \cite{giorelli2015nn}; the $410$\,mm assembled length of the I-Support arm is taken from Manti et al.\ \cite{manti2016soft}; the $100$\,mm workspace of Dataset 2 is the observed tip span after preprocessing (Section~\ref{sec:eda_findings}) and is used in lieu of a documented arm length. Entries marked ``---'' indicate metrics that the source did not report in directly comparable form.}
\label{tab:disc_benchmarks}
\begin{tabularx}{\linewidth}{@{}l X X X X@{}}
\toprule
Source                                              & Platform                      & Method                      & Mean error (mm)   & Relative (\%\,of $L$) \\
\midrule
Giorelli et al.\ \cite{giorelli2015nn}               & $280$\,mm cable arm            & Analytical Jacobian          & $15.12$             & $5.40$                  \\
Giorelli et al.\ \cite{giorelli2015nn}               & $280$\,mm cable arm            & Feed-forward network         & $7.35$              & $2.62$                  \\
Fang et al.\ \cite{fang2022jacobian}                 & pneumatic soft robot           & Learned-Jacobian iteration   & sub-mm              & ---                     \\
This thesis, Dataset 1 (Section~\ref{sec:res_traj})   & $410$\,mm I-Support            & Jacobian-learned iteration   & $0.77$              & $0.19$                  \\
This thesis, Dataset 2 (Section~\ref{sec:res_traj})   & $\sim\!100$\,mm three-valve arm & Sim-to-real feed-forward     & $6.20$              & $6.20$                  \\
\bottomrule
\end{tabularx}
\end{table}

\paragraph{Gap to the Ma et al.\ benchmark.}
Ma et al.\ \cite{ma2022bp} reported a pressure-space MAPE below $5\%$ on a three-chamber pneumatic arm with $216$ training samples; the best-performing Dataset 2 configuration in this thesis reaches $34.4\%$ MAPE with $12{,}650$ training samples (Section~\ref{sec:res_mlp_tip}). The $6.9\times$ gap is not attributable to dataset size, since the thesis dataset is about sixty times larger than Ma's, and it is most likely attributable to three differences in the experimental conditions. First, Ma's arm was purpose-built for the experiment with documented actuation units (bar) and a documented straight-pose origin, whereas the Dataset 2 commands are integer-valued on an undocumented scale $[0, 2500]$ and the recorded origin sits near the straight-pose tip rather than the robot base (Section~\ref{sec:res_pcc_single}). Second, Ma's dataset was recorded in quasi-static sampling mode, whereas the Dataset 2 rows are IID random actuations (Section~\ref{sec:eda_findings}) in which stochastic hysteresis contributes directly to the residual. Third, the feed-forward network used here has no hysteresis term; the recurrent extension of Section~\ref{sec:res_lstm} shifted Dataset 2 MAPE from $34.5\%$ to $34.2\%$, a small improvement consistent with Dataset 2's weak temporal structure rather than with the absence of a physical hysteresis effect. A tighter comparison against Ma's benchmark would require either a recalibrated Dataset 2 in documented units or a Ma-style purpose-built recording with the random-walk protocol used here.

\paragraph{Feed-forward network beats the analytical baseline, as predicted by Giorelli.}
Giorelli et al.\ \cite{giorelli2015nn} compared a feed-forward inverse-statics network against an analytical Jacobian iteration on a $280$\,mm cable-driven soft arm and reported a $2.05\times$ accuracy advantage for the network ($7.35$\,mm mean error against $15.12$\,mm) together with a $2160\times$ speed advantage at inference. The Dataset 2 results of this thesis reproduce the ordering: the tip-only feed-forward network reached pressure MAE $150.1$ against $285.2$ on the reachable subset of the single-segment PCC baseline (Section~\ref{sec:res_mlp_tip}), a $1.9\times$ advantage that matches Giorelli's $2.0\times$ figure within reading accuracy. In position-space terms, the Dataset 2 trajectory-tracking error of $7.40$\,mm on an approximately $100$\,mm workspace corresponds to $7.4\%$ relative error against Giorelli's $2.62\%$ on their $280$\,mm arm. The difference in relative error is consistent with the undocumented-units and zero-frame issues above, and with the smaller absolute workspace of Dataset 2 placing any fixed measurement noise proportionally closer to the target.

\paragraph{Jacobian-learned IK reaches the sub-millimetre level reported by Fang et al.}
Fang et al.\ \cite{fang2022jacobian} introduced the learned-Jacobian inverse-kinematics scheme adopted in Section~\ref{sec:jacobian} and reported sub-millimetre mean tracking error on their pneumatic soft-robot platform. The Dataset 1 configuration of Section~\ref{sec:res_jacobian} reached a mean closed-loop position error of $0.71$\,mm on the per-sample test sub-sample and $0.77$\,mm on the reference trajectories of Section~\ref{sec:res_traj}, which is at the same order of magnitude as Fang's result on a different physical platform. This is one of the cleaner literature replications available in the thesis, and it suggests that the Fang method generalises to the random-walk pressure regime of Dataset 1 without special tuning beyond the reference hyperparameters reported by the original authors. The sim-to-real extension that is Fang's main contribution was not validated here: the negative transfer result of Section~\ref{sec:res_sim_transfer} shows only that a Manti-inspired rod does not serve as a useful pre-training prior for the undocumented Dataset 2 robot, and a different simulator whose workspace matches Dataset 2 could still recover Fang's sim-to-real benefit.

\paragraph{Limits of cross-platform comparison.}
None of the three reference works above uses exactly the same robot, dataset, or metric as this thesis, so the numerical comparisons are indicative rather than definitive. Ma's arm has a different geometry and known actuation units; Giorelli's arm is cable-driven and its task is planar; Fang's platform is a different pneumatic robot with documented calibration and sim-to-real infrastructure. The three reference numbers are therefore best read as scale markers for what is achievable on cleanly characterised soft-robot platforms, rather than as targets that the present thesis is expected to hit on Dataset 2. The quantitative claim that survives these caveats is the one supported by the cleaner Dataset 1 geometry: that the warm-started Jacobian-learned iteration reaches sub-millimetre closed-loop error on a real two-segment pneumatic recording, matching Fang's sub-millimetre order of magnitude and improving by roughly an order of magnitude against the analytical-Jacobian accuracy that Giorelli reported on a simpler platform.

\section{The role of mid-segment information on Dataset 1}
\label{sec:disc_mid}

\paragraph{Why tip-only IK is under-determined on a two-segment arm.}
The I-Support manipulator of Dataset 1 has six chamber pressures and a three-dimensional tip. Even under the piecewise-constant-curvature model, a given tip position can be reached by a continuous one-parameter family of per-segment configurations: the first segment can bend more and the second less, or the reverse, with the tip held fixed. The inverse is therefore not a function but a relation, and a supervised regressor trained with a squared-error loss will gravitate toward the training-distribution posterior mean over the valid configurations for each tip, as argued in Section~\ref{sec:disc_interpretation}. The tip-only feed-forward network of Section~\ref{sec:res_mlp_tip} plateaued at $R^{2} = 0.32$ after four training epochs, and two of its six per-chamber predictions sat at or below chance ($R^{2} = -0.03$ for chamber 2 and $0.04$ for chamber 4 in Fig~\ref{fig:res_mlp_pred_vs_actual}). Both observations are consistent with the network having learned what it can learn and the remaining variance belonging to the posterior itself rather than to model capacity.

\paragraph{Mid-marker input concentrates the posterior.}
Augmenting the input with the mid-marker coordinates adds three partial observations of the internal configuration of the arm---specifically, the centroid of the three mid-section markers (Section~\ref{sec:ds1}). In the PCC reading of this geometry those three coordinates identify the end of the first segment, which, together with the tip position, over-determines the per-segment bending angle, azimuth, and elongation to within optical marker noise. The feed-forward ablation of Section~\ref{sec:res_mlp_tipmid} measured the effect: pressure MAE fell from $20.13$ to $17.05$ (a $15\%$ reduction), $R^{2}$ rose from $0.32$ to $0.44$, and the Wilcoxon signed-rank test on the paired per-frame errors returned $p \approx 2.4 \times 10^{-211}$. For a change in input features with no change in architecture or training protocol (Section~\ref{sec:mlp_tipmid}), this is an unusually clean intervention: whatever improvement the mid marker provides is attributable to information content rather than to model capacity or training-loop differences.

\paragraph{Temporal context amplifies the mid-marker signal rather than replacing it.}
A notable asymmetry emerged in the recurrent experiments of Section~\ref{sec:res_lstm}. The tip-only LSTM was marginally worse than the tip-only feed-forward network on Dataset 1 (per-frame RMSE $24.88$ against $24.07$; $R^{2} = 0.28$ against $0.32$), indicating that a ten-frame trajectory of tip-only observations carries no additional information that a regressor can use. The same LSTM applied to tip-plus-mid input produced $R^{2} = 0.51$, a $17\%$ relative improvement over the feed-forward tip-plus-mid configuration. The most straightforward reading is that the recurrent network is extracting the trajectory of the internal configuration rather than the trajectory of the tip. Historical tip samples alone have no unique mapping to historical internal configuration and therefore remain under-determined; once the mid-marker coordinates pin the internal state at every frame, the history of that state becomes a physically meaningful feature whose temporal derivatives carry velocity and hysteresis information. Temporal modelling of soft-robot inverse kinematics is therefore not a universally useful architectural choice: its benefit is contingent on the input features already resolving the under-identification that would otherwise corrupt the time series.

\paragraph{Caveat.}
The mid-marker benefit is contingent on a real mid-marker signal at deployment: the trajectory evaluation of Section~\ref{sec:res_traj} substituted the training-distribution centroid and the network extrapolated off its training slice to $19.08$\,mm mean tracking error (the worst Dataset~1 entry in Table~\ref{tab:res_overall_ds1}), so any controller intending to use a tip-plus-mid network would need a running shape estimator of comparable quality.

\section{Temporal modelling and hysteresis}
\label{sec:disc_temporal}

\paragraph{Where temporal modelling helped, and where it did not.}
The recurrent networks of Section~\ref{sec:res_lstm} were predicted to capture hysteresis and other history-dependent effects that a feed-forward network cannot see. The prediction was borne out in only one of the three configurations tested. The tip-plus-mid recurrent network on Dataset 1 improved on its feed-forward counterpart by a modest but statistically clean margin ($R^{2} = 0.51$ against $0.44$). The tip-only recurrent network on Dataset 1 was marginally worse than the tip-only feed-forward network, and the tip-only recurrent network on Dataset 2 was marginally better ($4.8\%$ MAE reduction). On smooth reference trajectories the recurrent and feed-forward networks were indistinguishable within about a millimetre on every trajectory of either dataset (Section~\ref{sec:res_traj}). Temporal modelling therefore did not deliver the uniform benefit one would expect if hysteresis were the dominant residual error on these datasets.

\paragraph{Why Dataset 1 tip-only does not benefit from history.}
A $K$-frame window of past inputs only helps a regressor when those past inputs constrain the mapping from the current input to the current output. On the two-segment Dataset 1 arm, the mapping from tip position to chamber pressure is multi-valued (Section~\ref{sec:disc_mid}), and the tip velocity across the window does not identify which branch of the solution manifold the current configuration lies on, because the per-segment bending split is not observable from the tip alone. The recurrent network therefore receives a richer input with no accompanying reduction in under-identification, and its optimisation collapses to the same posterior mean that the feed-forward network already reached. The window-length sweep of Section~\ref{sec:res_lstm} exposed the symptom cleanly: the best-epoch index of the tip-only recurrent training run fell from $15$ at $K = 5$ to $1$ at $K = 20$, so longer windows accelerated over-fitting rather than recovering signal.

\paragraph{Why Dataset 1 tip-plus-mid does benefit.}
Once the mid-marker coordinates pin the internal state at every frame (Section~\ref{sec:disc_mid}), the $K$-frame window of those coordinates carries the velocity and acceleration of the internal state, and therefore the direction in which the arm is currently moving. For a hysteretic system the mapping from configuration to pressure depends on the recent history of configuration, not only on the current configuration, and the tip-plus-mid recurrent network is the only configuration tested that has access to the features it would need to read that history. The magnitude of the improvement ($R^{2}$ from $0.44$ to $0.51$) suggests that a hysteretic signal is indeed present in Dataset 1, though not enough to dominate the remaining residual error. A cleaner test would require either a dataset with a more structured temporal protocol than the random-walk excitation used here, or an explicit hysteresis parametrisation (for example a Preisach or Bouc-Wen model) fit to the recovered internal-state trajectory.

\paragraph{Pneumatic transients set the window length.}
Alessi et al.\ \cite{alessi2024pushing} Table~1 report asymmetric AM-I-Support transients ($0.2$\,s pressurisation / $0.56$\,s depressurisation), which at DS1's $20$\,Hz correspond to $\sim$$4$ / $\sim$$11$ frames---above the single-frame MLP's horizon but within $K \geq 10$ of the LSTM. Empirical cross-correlation on DS1 (Fig~\ref{fig:app_ds1_lag}) gives a median lag of $13$ frames with chamber~$5$ peaking at $0.25$\,s ($R = 0.87$), consistent with Alessi's pressurisation value to within one frame.

\paragraph{Why Dataset 2 benefits only marginally.}
Dataset 2 behaves as an IID sample set (Section~\ref{sec:eda_findings}); the step-size ratio of $1.00$ in valve-command space and $0.96$ in tip space leaves almost no temporal structure for a recurrent network to exploit. The $4.8\%$ MAE reduction from the recurrent model is unlikely to reflect captured hysteresis, since there is no temporal order in which hysteresis could manifest. A more plausible explanation is that windowing acts as a mild regulariser: adjacent rows with similar tip positions supply a local averaging signal that reduces per-sample prediction variance. The monotone benefit with longer $K$ (MAE $148.05 \to 138.68$ between $K = 5$ and $K = 20$) is consistent with a regularisation story rather than with a temporal-signal story. For a classifier on truly IID data longer windows would not be expected to help at all; the small positive gradient here is plausibly attributable to residual spatial adjacency between neighbouring rows in the recording order rather than to time.

\paragraph{Smoothness aside.}
On the reference trajectories of Section~\ref{sec:res_traj} the recurrent and feed-forward networks were within one millimetre of each other in tracking error, but the LSTM's pressure-command jerk was one to two orders of magnitude lower on Dataset~1---a property of the hidden-state low-pass, useful where actuator wear or valve chatter matters.

\section{Jacobian-learned iteration versus direct regression}
\label{sec:disc_jacobian_vs_mlp}

\paragraph{Recap: position-space winner versus pressure-space winner.}
The Fang 2022 trade-off summarised in Section~\ref{sec:disc_interpretation} gives one global statement of the choice between the two method families: a direct-regression network is the posterior-mean estimator over valid pressure solutions and therefore minimises pressure-space squared error, while a Jacobian iteration commits to a single posterior-mode solution near its initialisation and therefore minimises position-space error at inference. The present section decomposes that global trade-off into four operational axes---generalisation, sample efficiency, out-of-distribution robustness, and runtime---on which the choice has to be made in practice.

\paragraph{Generalisation is bounded by different things.}
A direct-regression network's generalisation to novel tip targets is bounded by the coverage of the training distribution in tip space: any target outside the convex hull of training tips is extrapolation, and the network's prediction has no formal guarantee. A Jacobian iteration's generalisation is bounded instead by the accuracy of the forward-kinematics network on the iterates that the solver visits during the solve. Two consequences follow. First, the Jacobian method inherits the failure modes of the forward network: a forward network with test-set MAE of $2.55$\,mm on Dataset 2 (Section~\ref{sec:res_jacobian}) caps the Jacobian iteration's achievable closed-loop error at roughly that scale, regardless of how many iterations are spent. Second, the Jacobian method can in principle reach targets outside the inverse-kinematics training distribution, because its search is anchored by the forward model rather than by a learned inverse map. The sub-millimetre result on the Dataset 1 per-sample test sub-sample---where the seed-0 test trajectories were never seen at training time---is a partial demonstration: the warm-started iteration hit the tolerance floor on every target in the sub-sample (Table~\ref{tab:res_jacobian_pos}).

\paragraph{Sample efficiency favours the Jacobian formulation in principle.}
A direct-regression inverse-kinematics network has to learn a map across the full target distribution, and its sample complexity scales with the complexity of that distribution. A Jacobian iteration only needs its forward-kinematics network to fit the pressure-to-position map, which is single-valued regardless of actuation redundancy and is therefore cheaper to learn: the Dataset 1 forward network reached test MAE $12.97$\,mm after only two epochs (Section~\ref{sec:res_jacobian}), against the four epochs required by the tip-only inverse network of Section~\ref{sec:res_mlp_tip}. In the limit of small real-data budgets this argument suggests a Jacobian iteration would remain usable where a direct-regression network over-fits; a systematic data-efficiency sweep of the Jacobian method on a downsampled real-data budget is a natural extension of the sim-to-real study of Section~\ref{sec:res_sim_transfer} and would quantify the above intuition.

\paragraph{Out-of-distribution robustness is the honest cost of the Jacobian iteration.}
The clearest out-of-distribution example in the thesis is the Dataset 2 helical spiral of Section~\ref{sec:res_traj}: its targets trace the full $z$-half-extent of the observed workspace and include configurations close to the edge of what the forward-kinematics network represents reliably. The warm-started Jacobian iteration reached a mean tracking error of $31.94$\,mm on the spiral, with a worst-case single-target error of $258$\,mm, compared with $7.42$\,mm and $8.17$\,mm for the sim-to-real and tip-only feed-forward networks on the same trajectory. A single diverging DLS run on a three-valve arm is enough to invalidate the mean because there is no actuation redundancy available to recover from a poor update; feed-forward methods cannot exhibit this behaviour because a single forward pass cannot diverge. Operationally, this means the Jacobian iteration's sub-millimetre advantage is conditional on staying inside the workspace where the forward-kinematics network is well-conditioned, and it will fail gracefully only when explicit workspace bounds or iteration-based safety checks are added to the solver.

\paragraph{Runtime argues against the Jacobian iteration at high control rates.}
The warm-started Jacobian iteration cost about $46$\,ms per target (Section~\ref{sec:res_jacobian}) against well under $1$\,ms for the feed-forward networks, two to three orders of magnitude slower depending on the measurement context. Both sit inside a $50$\,ms budget appropriate for a $20$\,Hz controller, but neither the Jacobian method nor any other method tested here would meet a $1$\,kHz control loop on the available hardware. The bulk of the Jacobian cost is the reverse-mode automatic differentiation of the forward network and the linear solve at each step (Eq.~\eqref{eq:dls_step}); runtime therefore scales linearly with the number of iterations and with the width of the forward network, not with the dataset size or the target's distance from the training distribution. The implication for practical soft-robot controllers is regime-dependent. For quasi-static pick-and-place tasks the Jacobian iteration's accuracy gain is the dominant consideration and the latency is irrelevant. For dynamic tasks that require closed-loop updates at kilohertz rates the feed-forward methods remain the only candidate unless the forward-network evaluation can be specialised (for example, to a fixed configuration neighbourhood in which a Taylor expansion replaces the autograd Jacobian).

\paragraph{The hybrid is already the default.}
The warm-started configuration adopted for every headline Jacobian number in Chapter~\ref{chap:results} is itself a hybrid: the direct-regression network's prediction supplies the initial pressure $\mathbf{p}_0$, and the DLS iteration then refines it. On Dataset 2, warm-starting reduced the mean iteration count from $16.8$ (zero initialisation) to $2.3$ and lifted the convergence rate from $78\%$ to $99\%$ (Section~\ref{sec:res_jacobian}). The hybrid inherits the direct network's broad-coverage generalisation from the zero initialisation and the Jacobian iteration's sub-millimetre refinement at inference. Three natural extensions of this hybrid would test its limits: a nearest-neighbour initialisation from a cached training distribution, a learned stopping criterion that exits the iteration before divergence, and a two-stage controller that runs the feed-forward network in a fast inner loop and invokes the Jacobian iteration only when the position error exceeds a threshold. None of these were tested in the present thesis and all are compatible with the architectures and checkpoints of Chapter~\ref{chap:methods}.

\section{Sim-to-real observations}
\label{sec:disc_sim}

\paragraph{Summary.}
Strategy~B matched Strategy~A within $\pm 1$ command unit at every budget; Strategy~C reached $R^{2} = -1.71$. Two findings, one diagnosis: the reality gap here is dominated by geometric workspace mismatch, not by sample count or material fidelity.

\paragraph{Workspace geometry dominates the reality gap.}
The synthetic tip cloud spans $\approx 500 \times 480 \times 330$\,mm against Dataset~2's $102 \times 104 \times 64$\,mm; the two distributions essentially do not overlap. Other axes---hyperelasticity, viscoelastic hysteresis, pneumatic lag (absent via Eq.~\eqref{eq:sim_moment}), manufacturing asymmetry---contribute in principle, but Strategy~C's $R^{2} = -1.71$ is set by the coordinate mismatch: a network trained in the wrong region cannot map real inputs to the right output regardless of material fidelity. Strategy~B's fine-tuning then discards the pre-training because the synthetic input range has no counterpart in the real distribution (a $(150,-200,-300)$\,mm synthetic tip has no Dataset~2 analogue).

\paragraph{Why Fang et al.\ succeeded.}
Fang et al.\ \cite{fang2022jacobian} used a digital twin of the target platform. The simulator here was a Manti-inspired reference rod because Dataset~2's parameters are undocumented (Section~\ref{sec:res_pcc_single}). The negative result is therefore an argument for Fang's method \emph{conditional on a digital twin}, not a refutation.

\paragraph{Minimum changes for a positive transfer.}
(1)~Rescale and re-centre the simulator workspace to the target; this removes the dominant axis. (2)~Calibrate pressure-to-command scale so synthetic commands cover the recorded range. (3)~Add a hysteresis term (e.g.\ a Preisach layer). The first is most tractable; the other two address axes the present design could not isolate.

\section{Limitations}
\label{sec:disc_limitations}

\subsection{No physical robot}
\label{sec:disc_lim_hw}

The sub-millimetre Jacobian result on Dataset 1 and the $6.20$\,mm sim-to-real tracking on Dataset 2 were obtained by running every method's predicted pressure through the shared learned forward-kinematics arbiter of Section~\ref{sec:res_jacobian}. The arbiter is itself an approximation of the real robot, and its own test-set MAE of $2.55$\,mm on Dataset 2 and $12.97$\,mm on Dataset 1 sets an upper bound on the accuracy the closed-loop evaluations can distinguish between methods. No closed-loop control experiment was run on physical hardware in this thesis. Three consequences follow. First, controller latency, sensor noise, and actuator saturation are not tested; the reported inference times (Section~\ref{sec:res_traj}) exclude the overhead of sensor reading and valve command transmission on a real controller. Second, the Dataset 2 spiral divergence of the Jacobian iteration (Section~\ref{sec:res_traj}) is a failure of the arbiter's reliability at workspace edges, and it cannot be determined from the present data whether a real-hardware run would exhibit the same divergence or would stabilise under closed-loop feedback from the real tip position. Third, any claim about real-world tracking quality on these platforms is an expectation derived from a learned model rather than a measured performance. Closing this limitation requires access to the physical Dataset 1 or Dataset 2 arm, or a new physical pneumatic platform on which the methods can be retrained from scratch.

\subsection{Dataset coverage}
\label{sec:disc_lim_data}

The conclusions of this thesis are drawn from two datasets, both on pneumatic soft manipulators, both with sensing by optical markers, and both in a workspace of at most a few hundred millimetres. Three kinds of generalisation are therefore not tested here. First, cable-driven and tendon-driven soft arms---including the platform studied by Giorelli et al.\ \cite{giorelli2015nn}---have different actuation dynamics, different hysteresis mechanisms, and different failure modes; the mid-marker argument of Section~\ref{sec:disc_mid} may or may not transfer to them. Second, larger-scale manipulators (in the metre range, for example surgical or industrial soft manipulators) have different noise-to-workspace ratios, and the sub-millimetre Jacobian claim could degrade on platforms where optical noise is a larger fraction of the target accuracy. Third, many engineering soft arms use full-shape sensing---fibre-Bragg-grating fibres, proprioceptive inertial cells, or capacitive skin---rather than optical markers at discrete triads; the specific mid-marker ablation of Section~\ref{sec:res_mlp_tipmid} may not translate directly when the shape signal is continuous along the arm rather than sampled at the inter-module joint. The three-property decision framework of Section~\ref{sec:disc_interpretation} (actuation redundancy, sensing coverage, target-domain overlap) was drawn from $n = 2$ datasets, and widening the dataset pool is the principal direction in which its claims remain to be verified.

\subsection{Simulator fidelity}
\label{sec:disc_lim_sim}

The PyElastica simulator of Section~\ref{sec:sim_setup} is a single-module Cosserat rod with a linear elastic constitutive law, no intrinsic viscoelasticity, instantaneous pressure application as a bending moment, and no manufacturing imperfections. The five reality-gap axes identified in Section~\ref{sec:disc_sim}---workspace geometry, constitutive law, viscoelasticity, air-flow dynamics, and manufacturing asymmetries---were acknowledged, but only the workspace-geometry axis was measured quantitatively. Three consequences follow. First, the negative sim-to-real result on Dataset 2 is consistent with any combination of the five gaps contributing to the residual, and the present data cannot decompose the $150$-command-unit Strategy-A MAE into per-axis contributions. Second, the Young's-modulus calibration of Section~\ref{sec:res_sim_cal} was done against I-Support anchors from Manti et al.\ \cite{manti2016soft} and not against Dataset 2 data, because Dataset 2's physical parameters are undocumented; the calibrated rod is an I-Support-like reference, not a model of the Dataset 2 robot. Third, any quantitative claim about the value of a better simulator is necessarily hypothetical under the present experimental design. A follow-up simulator that combines a hyperelastic material model (Ecoflex 00-50 or similar), viscoelastic damping, and a simple pneumatic line-delay term would be the minimal platform on which the per-axis attribution could be measured.

\subsection{Metric choice and statistical evidence}
\label{sec:disc_lim_metric}

All position-space errors in this thesis use the tip position only. Two configurations with the same tip position but different per-segment bending or mid-section location are scored as having zero relative error, even if the intermediate shape of the arm differs by several centimetres. For applications that require obstacle avoidance, full-arm contact prediction, or collaborative manipulation, a full-shape error metric---such as the Hausdorff distance between predicted and achieved arm centrelines---would be more appropriate than the tip-only Euclidean error used here. The pressure-space metrics (MAE, MAPE, NMAE, $R^{2}$) reported throughout Chapter~\ref{chap:results} also carry the caveats discussed in Section~\ref{sec:res_jacobian}: on multi-valued inverses they favour posterior-mean estimators and therefore do not rank methods by any single notion of ``quality''.

A second limitation of the reported evidence is statistical. Each network reported in Chapter~\ref{chap:results} was trained once from a single random seed (Section~\ref{sec:hyperparam}). The paired Wilcoxon signed-rank test and bootstrap confidence interval reported for the tip-plus-mid ablation in Section~\ref{sec:res_mlp_tipmid} are the only uncertainty quantifications in this thesis, and they apply only to the comparison between two configurations on identical test frames. For comparisons across different methods or different training runs the variance across seeds is not bounded. A follow-up study in which each configuration is trained three to five times from independent seeds, with confidence intervals reported on every summary metric, would tighten the claims of Sections~\ref{sec:disc_jacobian_vs_mlp} and~\ref{sec:overall} about which method is best in each regime. The cost of such a study---roughly tripling the training compute---is manageable within the Chapter~\ref{chap:methods} infrastructure and is the cheapest intervention that would strengthen the evidence base.

\section{Implications}
\label{sec:disc_implications}

This section collects the practical implications of Chapter~\ref{chap:results} for four audiences: practitioners selecting an inverse-kinematics method for a new pneumatic soft-robot arm, hardware and sensor designers specifying new platforms, simulation engineers building sim-to-real pipelines, and control engineers deploying soft-robot controllers.

\paragraph{A checklist for method selection.}
The three-property framework of Section~\ref{sec:disc_interpretation} gives a compact decision criterion that a practitioner can evaluate before writing any inverse-kinematics code. First, count the actuation degrees of freedom and compare them with the output degrees of freedom of the controlled variable. If actuators outnumber outputs, the inverse is multi-valued and a Jacobian-learned iteration can be expected to reach closed-loop position accuracy at the tolerance floor (Section~\ref{sec:res_jacobian}), at the cost of roughly two orders of magnitude in per-target compute. If actuators match outputs, the inverse is essentially single-valued and a direct-regression network will be the simpler and faster choice with comparable accuracy. Second, assess what partial observations of the arm's internal configuration are available at deployment; the tip alone is generally insufficient on multi-segment arms (Section~\ref{sec:disc_mid}), and a single additional triad at an inter-module interface is both cheaper and more informative than a larger network. Third, if a simulator is available, measure the overlap between its synthetic workspace and the intended real operating region \emph{before} running any pre-training; if the overlap is small, as it was on Dataset 2, the sim prior will not help even at aggressive fine-tuning budgets (Section~\ref{sec:disc_sim}).

\paragraph{Implications for hardware and sensor design.}
The headline quantitative result on Dataset 1---pressure-space $R^{2}$ rising from $0.32$ to $0.51$ when a single mid-marker triad is added to the network input---is an argument for investing in partial-shape sensing rather than in deeper networks or larger datasets on multi-segment soft arms. A practitioner designing a new two-module pneumatic arm would do more to improve inverse-kinematics accuracy by embedding one row of markers at the inter-module joint than by switching to any of the more complex networks tested in this thesis. A second hardware implication follows from the Dataset 2 experience of unknown actuation units and an undocumented zero-frame: the $6.9 \times$ pressure-MAPE gap to Ma et al.\ \cite{ma2022bp} (Section~\ref{sec:disc_benchmarks}) was not a method-quality gap but a calibration-metadata gap, and it could have been closed by recording a calibration protocol alongside the raw data. New datasets should document actuation units, the physical straight-pose origin, and at least a small held-out calibration sequence that future methods can reinterpret. Finally, for arms intended for closed-loop deployment rather than offline evaluation, a continuous shape sensor such as fibre-Bragg-grating fibre or embedded proprioceptive cells would remove the deployment-time out-of-distribution risk that Section~\ref{sec:disc_mid} identified for the tip-plus-mid network: the training-distribution mid-marker centroid had to be substituted at inference for the trajectory evaluation, and the network extrapolated off its training slice to $19.08$\,mm error.

\paragraph{Implications for simulation and sim-to-real practice.}
The negative transfer result of Section~\ref{sec:res_sim_transfer} suggests a reordering of priorities when building a simulator for a pneumatic soft robot. The single change most likely to convert a Dataset 2-style negative transfer into a positive one is not a richer material model or a larger synthetic dataset but a workspace rescaling that makes the synthetic tip distribution overlap the real one (Section~\ref{sec:disc_sim}). Rescaling is close to free; hyperelastic material models and viscoelastic damping are expensive to implement and to calibrate. A pragmatic pipeline therefore starts with a workspace-matched simulator from the outset and adds per-axis fidelity only once the workspace-aligned version has been shown to help. For an undocumented real arm on which a digital twin cannot be built, the honest conclusion is that sim-to-real is unlikely to deliver value and that any compute budget is better spent on real-data recording; this is the reading that Section~\ref{sec:disc_sim} gave of the contrast between Fang et al.\ \cite{fang2022jacobian}'s positive sim-to-real result and the present negative one.

\paragraph{Implications for control engineering.}
A control engineer deploying any of the methods in this thesis should choose the method that matches the downstream metric. If the controlled quantity is the tip position under closed-loop feedback, the warm-started Jacobian-learned iteration is the right default (Section~\ref{sec:disc_jacobian_vs_mlp}); if the controlled quantity is the actuation command and the downstream consumer cares about pressure-space MAPE, the direct-regression network is a better fit and has sufficient latency headroom to meet a kilohertz loop. The hybrid of a feed-forward warm start followed by Jacobian refinement---which this thesis already uses by default---is both faster and more accurate than either extreme; a two-stage controller that runs the feed-forward network in the inner loop and triggers the Jacobian iteration only when the measured position error exceeds a threshold would combine the latency of the first method with the accuracy of the second. On smooth target trajectories the recurrent network is worth keeping even at equivalent accuracy: its pressure-command jerk is one to two orders of magnitude lower than that of the feed-forward methods (Section~\ref{sec:res_traj}), and the reduced jerk translates directly into lower actuator wear and lower energy consumption.

\paragraph{Implications for the broader soft-robotics research programme.}
The overall shape of the results supports three claims about the direction of soft-robot inverse-kinematics research. First, classical piecewise-constant-curvature inverse kinematics is not competitive with small feed-forward networks on either of the two real datasets tested here, for two different reasons---a constant-arc-length structural ceiling on Dataset 2 and a multi-valued-inverse calibration collapse on Dataset 1 (Sections~\ref{sec:res_pcc_single}, \ref{sec:res_pcc_multi}). PCC remains a useful analytical reference, not a deployable method. Second, the largest single change this thesis achieved on a learned inverse-kinematics pipeline came from three marker coordinates rather than from any network architecture change; partial-shape sensing on multi-segment arms is therefore a more impactful research direction than deeper or wider networks. Third, sim-to-real transfer on a pneumatic soft robot requires a simulator whose geometry is aligned with the target robot: a biologically plausible or architecturally similar rod is not a drop-in prior on an undocumented platform, and the community is best served by treating workspace-matched digital twins as a prerequisite for the sim-to-real strategy rather than as a convenience. These three claims are the compact summary that this thesis would offer to a soft-robotics reader deciding where to invest their next experiment.
