\chapter{Conclusion}
\label{chap:conclusion}

This thesis evaluated four inverse-kinematics method families---analytical piecewise-constant-curvature, feed-forward neural, recurrent neural, and Jacobian-learned iteration---on two pneumatic soft-robot datasets (a six-chamber I-Support recording and an undocumented three-valve recording) under identical training and evaluation protocols, complemented by a four-real-data-budget sim-to-real study and a four-trajectory tracking evaluation.

The analytical baseline is not competitive with even a small feed-forward network on either dataset, for two distinct reasons: a constant-arc-length ceiling gives a $22$\,mm residual on Dataset~2, and a multi-valued-inverse calibration collapse gives a $51$\,mm closed-loop error on Dataset~1 despite a near-zero geometric residual. Adding the mid-marker input raised the Dataset~1 feed-forward $R^{2}$ from $0.32$ to $0.44$, and the recurrent tip-plus-mid configuration reached $R^{2} = 0.51$ (the best Dataset~1 result here). The warm-started Jacobian-learned iteration reached $0.71$\,mm per-sample and $0.77$\,mm trajectory closed-loop error on Dataset~1 but diverged on the Dataset~2 spiral; on Dataset~2 the Strategy-B sim-to-real feed-forward network was the best tracker at $6.20$\,mm. Sim-to-real pre-training gave no measurable improvement at any real-data budget, attributable to a fivefold workspace mismatch rather than the sample count. Against published benchmarks, the Dataset~1 sub-mm Jacobian result matches Fang et al.\ \cite{fang2022jacobian} and the feed-forward-over-PCC improvement on Dataset~2 matches the $\sim$$2\times$ pattern of Giorelli et al.\ \cite{giorelli2015nn}; the sub-$5\%$ MAPE of Ma et al.\ \cite{ma2022bp} was not reached on Dataset~2, with most of the gap attributable to undocumented actuation units rather than method quality.

The synthesis is a three-property decision criterion: method choice depends on actuation redundancy, sensing coverage, and the overlap between any simulator prior and the real operating workspace. Jacobian-IK dominates where redundancy exists and the FK arbiter is well-conditioned; feed-forward (with or without sim-to-real pre-training) dominates where actuation matches the output dimension and no digital twin is available; partial-shape sensing beats architectural depth when the inverse is under-identified.

These findings operationalise \emph{AI for Control of Soft Robot by Using ML and Inverse Kinematics} per §\ref{sec:intro_scope}: \emph{AI/ML} as the four learning-based families; \emph{Inverse Kinematics} as the analytical PCC baseline together with the Jacobian-learned hybrid; \emph{Control} in its open-loop position-control sense, validated by four-trajectory tracking. Closed-loop real-hardware validation remains the natural next step (Chapter~\ref{chap:future_work}).
