\chapter{Future Work}
\label{chap:future_work}

Five directions follow directly from the limitations of Chapter~\ref{chap:discussion} and would strengthen or extend the three-property decision framework.

\paragraph{Closed-loop hardware validation.}
\label{sec:fw_hw}
Every closed-loop number here is arbiter-mediated (FK MAE $2.55$ / $12.97$\,mm caps between-method resolution). Re-deploying the best DS1 (warm-start Jacobian-IK) and DS2 (Sim2Real FF) configurations on physical hardware would measure the true tracking error under real sensor noise, latency, and actuator saturation, and would confirm or overturn the arbiter-space spiral divergence.

\paragraph{Workspace-matched simulators.}
\label{sec:fw_sim}
The tractable first step is to rescale and re-centre the PyElastica workspace to Dataset~2's tip distribution and re-run Strategies~A/B/C at the four budgets. If the $B - A$ gap remains within noise, adding hyperelasticity, viscoelastic damping, and air-flow delay one axis at a time would isolate per-axis contributions to the reality gap.

\paragraph{Broader sensing modalities and platforms.}
\label{sec:fw_sensing}
Three external-validity tests: (i) rerun the mid-marker ablation with continuous shape sensors (fibre-Bragg, proprioceptive cells) to check whether the $R^{2}$ $0.32 \to 0.51$ lift transfers to distributed sensing; (ii) evaluate the four IK families on a cable- or tendon-driven arm whose hysteresis mechanism differs from pneumatic; (iii) benchmark on a metre-scale arm where noise-to-workspace ratio erodes the sub-mm Jacobian advantage.

\paragraph{Multi-seed statistical evidence.}
\label{sec:fw_stats}
Retraining every configuration from $3$--$5$ independent seeds and reporting confidence intervals on every summary metric would tighten the cross-method rankings and the $0.01$--$0.05$\,mm Jacobian-initialisation differences. A three-seed rerun is a day of compute on the reported workstation.

\paragraph{Hybrid controllers and explicit hysteresis.}
\label{sec:fw_hybrids}
A two-stage controller (FF inner loop, Jacobian on error-threshold exceedance) would combine FF latency with Jacobian accuracy; a learned stopping criterion would address the DS2 spiral divergence; a Preisach or Bouc-Wen hysteresis layer between command and forward network would provide a physically grounded test of the hysteresis hypothesis that recurrent modelling could only approach statistically.
