\chapter{Abstract}
\thispagestyle{plain}

\noindent This thesis studies how AI-based methods compute actuation commands that drive a pneumatic soft manipulator's tip to a desired position---the inverse-kinematics (IK) problem, which is the standard enabling technology for open-loop position control of a continuum arm. Following the usage of Giorelli et al.\ \cite{giorelli2015nn}, Ma et al.\ \cite{ma2022bp}, Fang et al.\ \cite{fang2022jacobian}, and Centurelli et al.\ \cite{centurelli2022closed}, the title keywords are instantiated as: \emph{AI / ML} = four learning-based families (feed-forward, recurrent, Jacobian-learned iteration, sim-to-real fine-tuning); \emph{Inverse Kinematics} = classical piecewise-constant-curvature (PCC) inversion plus the Jacobian-learned hybrid; \emph{Control} = open-loop tip-to-pressure mapping, validated by four-trajectory tracking through a shared learned forward-kinematics arbiter. No closed-loop experiments on physical hardware are reported.

\noindent\textbf{Methods.}
The four IK families are compared head-to-head on two pneumatic datasets with complementary actuation dimensionality: a six-chamber two-module I-Support recording (Dataset~1) and an undocumented three-valve recording (Dataset~2). A paired mid-marker ablation quantifies partial-shape sensing on the over-actuated platform; a PyElastica sim-to-real study evaluates pre-training at four real-data budgets; a four-trajectory tracking protocol applies an identical evaluation to every method.

\noindent\textbf{Results.}
Classical PCC is not competitive with a small feed-forward network on either dataset: a $22$\,mm residual on Dataset~2 (constant-$L$ ceiling) and a $51$\,mm closed-loop error on Dataset~1 (multi-valued-inverse calibration collapse) despite a near-zero geometric residual. Adding mid-marker input raises the Dataset~1 pressure-space $R^{2}$ from $0.32$ to $0.51$, the largest single-factor improvement. The warm-started Jacobian-learned iteration reaches $0.77$\,mm mean tracking error on Dataset~1, matching Fang et al.\ \cite{fang2022jacobian}. Sim-to-real pre-training gives no measurable benefit on Dataset~2 at any budget, a negative result attributable to a fivefold synthetic--real workspace mismatch rather than sample count.

\noindent\textbf{Conclusions.}
No single method dominates every regime. Inverse-kinematics method choice for pneumatic soft robots depends on three physical properties of the platform and task---actuation redundancy, sensing coverage, and simulator--target workspace overlap---and this three-property criterion organises the regime-specific findings into a compact method-selection framework.
